Neural Power Units

### 2020-06-02 papers/main.tex
hyperparams

One of the most important properties of a layer in a neural network is its
ability to scale. With the large scale arithmetic task we show that the NPU
works reliably on many-input tasks that are heavily over-parametrized.  In this
section we compare NALU, NMU, NPU, and the NaiveNPU on a task that is identical
to the `arithmetic task' that \cite{madsen_neural_2020} and
\cite{trask_neural_2018} analyse as well.
The goal is to sum two subsets of a 100 dimensional vector and
apply an operation (like $\times$) to the two summed subsets.  The dataset
generation is defined in the set of Eq.~\ref{eq:large_scale_def}, with the
parameters from Tab.~\ref{tab:dataset_params_large_scale}.
\begin{align}
  \label{eq:large_scale_def}
  a = \sum_{i=s_{1,\text{start}}}^{s_{1,\text{end}}} x_i, &&
  b = \sum_{i=s_{2,\text{start}}}^{s_{2,\text{end}}} x_i, &&
  t_{\text{add}} = a + b, && t_{\text{mul}} = a \times b, &&
  t_{\text{div}} = 1 / a, && t_{\text{sqrt}} = \sqrt{a},
\end{align}
where starting and
ending values $s_{i,\text{start}},s_{i,\text{end}}$ of the summations are chosen
such that $a$ and $b$ come from subsets of the input vector $\bm x$ with a
given overlap.  Specifics of the used models are defined in
Tab.~\ref{tab:models_large_task}.  \cite{madsen_neural_2020} perform an
extensive analysis of this task with different subset and overlap ratios,
varying model and input sizes, and much more, establishing that the combination
of NAU/NMU outperforms the NALU. We focus on the
comparison of NPU, NMU, and NALU on the default parameters of
\cite{madsen_neural_2020} which sets the subset ratio to 0.5 and the overlap
ratio to 0.25 (details in Tab.~\ref{tab:dataset_params_large_scale}).
We also compare to the NaiveNPU (without the relevance gate) to show
how important the gating mechanism is for both sparsity and overall
performance.

Fig.~\ref{fig:pareto} plots training and testing errors over the number of
non-zero parameters for all models and tasks.
The addition column shows that all models except NaiveNPU learn ($+$) on the
training set, with the NMU converging to the sparsest models. The NMU also
performs best on the testing set, with the NPU as a close second.\\
On ($\times$), the best NMU models also outperform the NPU, but some of
them do not converge at all, which is why the NPU has a better testing MSE on
average (see also Tab.~\ref{tab:arithmetic100_val}). The testing MSE of the
NALU is so large that is excluded from the plot.\\
On ($\div,\sqrt{\cdot}$) the NPU clearly outperforms all other layers
in MSE and sparsity. Generally, the difference between the NaiveNPU and the NPU
is very significant and demonstrates how important the relevance gate is both
for convergence and sparsity. The NPU effectively converts irrelevant inputs
to 1s, while the NaiveNPU is stuck on the zero gradient plateau.

@@

We introduced the \emph{Neural Arithmetic Unit} that addresses the deficiencies
of current arithmetic units: it can learn arbitrary power functions for
positive, negative, and small numbers. We showed that the NPU outperforms its
main competitor (NALU) and reaches performance that is on par with the
multiplication specialist NMU.  For tasks that are known to contain only
addition and multiplication a stack of NAU and NMU can be a better choice than
the NPU, because they might be easier to train. For more complicated problems
that involve division or power functions, the NPU is the better choice.

Additionally we have demonstrated how the NPU can be used as an interpretable
model that can recover governing equations of dynamical systems purely from
data.

@@

\appendix
\section*{Appendix}%
\label{sec:appendix}

@@

\begin{table}[h]
  \centering
  \caption{Validation error of the different models (i.e. mean of each heatmap
  in Fig.~\ref{fig:simple_err}). Each value is obtained by averaging the error
  of 20 models.
  }
  \label{tab:simple_err}
  \input{../plots/simple_err.tex}
\end{table}

\begin{table}[h]
  \centering
  \caption{Model definitions for the simple arithmetic task}
  \label{tab:models_simple_task}
  \begin{tabular}{llll}
    \toprule
    Model & Layer 1 & Layer 2 & Layer 3 \\
    \midrule
    NPU & NAU(2,\,6) & NPU(6,\,2) & -- \\
    NMU & NAU(2,\,6) & NMU(6,\,2) & -- \\
    NALU & NALU(2,\,6) & NALU(6,\,2) & -- \\
    iNALU & iNALU(2,\,6) & iNALU(6,\,2) & -- \\
    Dense & Dense(2,\,10,\,$\sigma$) & Dense(10,\,10,\,$\sigma$) & Dense(10,\,2) \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[h]
  \caption{Model definitions for the large scale arithmetic task}
  \label{tab:models_large_task}
  \begin{tabular}{llll}
    \toprule
    Model & Layer 1 & Layer 2\\
    \midrule
    NPU & NAU(100,\,100) & NPU(100,\,1) \\
    NPU & NAU(100,\,100) & NPU(100,\,1) \\
    NMU & NAU(100,\,100) & NMU(100,\,1) \\
    NALU & NALU(100,\,100) & NALU(100,\,1) \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{table}[h]
  \centering
  \caption{fSIR model definitions}
  \label{tab:fsir_models}
  \begin{tabular}{llll}
    \toprule
    Model & Layer 1 & Layer 2 & Layer 3 \\
    \midrule
    NPU & $\npu(3,\,h)$ & $\nau(h,\,3)$ & -- \\
    NPU & NPU$_{\text{real}}(3,\,h)$ & $\nau(h,\,3)$ & -- \\
    Dense & $\dense(2,\,h,\sigma)$ & $\dense(h,\,h,\,\sigma)$ & $\dense(h,\,3)$ \\
    \bottomrule
  \end{tabular}
\end{table}

>>>>

One of the most important properties of a layer in a neural network is its
ability to scale. With the large scale arithmetic task we show that the NPU
works reliably on many-input tasks that are heavily over-parametrized.  In this
section we compare NALU, NMU, NPU, and the NaiveNPU on a task that is identical
to the `arithmetic task' that \cite{madsen_neural_2020} and
\cite{trask_neural_2018} analyse as well.
The goal is to sum two subsets of a 100 dimensional vector and
apply an operation (like $\times$) to the two summed subsets.  The dataset
generation is defined in the set of Eq.~\ref{eq:large_scale_def}, with the
parameters from Tab.~\ref{tab:dataset_params_large_scale}.
\begin{align}
  \label{eq:large_scale_def}
  a = \sum_{i=s_{1,\text{start}}}^{s_{1,\text{end}}} x_i, &&
  b = \sum_{i=s_{2,\text{start}}}^{s_{2,\text{end}}} x_i, &&
  t_{\text{add}} = a + b, && t_{\text{mul}} = a \times b, &&
  t_{\text{div}} = 1 / a, && t_{\text{sqrt}} = \sqrt{a},
\end{align}
where starting and ending values $s_{i,\text{start}},s_{i,\text{end}}$ of the
summations are chosen such that $a$ and $b$ come from subsets of the input
vector $\bm x$ with a given overlap.  The training objective is standard MSE,
regularized with L1:
\begin{equation}
  \mathcal{L} = \text{MSE}(\text{model}(\bm x), t) + \beta||\bm\theta||_1,
\end{equation}
where $\beta$ is scheduled to be low in the beginning of training and stronger
towards the end.  Specifics of the used models and their hyper-parameters are defined in
Tab.~\ref{tab:models_large_task} \& \ref{tab:training_large_scale}.  \cite{madsen_neural_2020} perform an
extensive analysis of this task with different subset and overlap ratios,
varying model and input sizes, and much more, establishing that the combination
of NAU/NMU outperforms the NALU. We focus on the
comparison of NPU, NMU, and NALU on the default parameters of
\cite{madsen_neural_2020} which sets the subset ratio to 0.5 and the overlap
ratio to 0.25 (details in Tab.~\ref{tab:dataset_params_large_scale}).
We also compare to the NaiveNPU (without the relevance gate) to show
how important the gating mechanism is for both sparsity and overall
performance.

Fig.~\ref{fig:pareto} plots testing errors over the number of non-zero
parameters for all models and tasks.  The addition plot shows only NMU and NPU
can extrapolate on ($+$) with the NMU converging to the sparsest models.\\
On ($\times$), the best NMU models outperform the NPU, but some of them do not
converge at all, which is why the NPU has a better testing MSE on average (see
also Tab.~\ref{tab:arithmetic100_val}). The testing MSE of the NALU is so large
that is excluded from the plot.\\
On ($\div,\sqrt{\cdot}$) the NPU clearly outperforms all other layers
in MSE and sparsity. Generally, the difference between the NaiveNPU and the NPU
is very significant and demonstrates how important the relevance gate is both
for convergence and sparsity. The NPU effectively converts irrelevant inputs
to 1s, while the NaiveNPU is stuck on the zero gradient plateau.

@@

We introduced the \emph{Neural Arithmetic Unit} that addresses the deficiencies
of current arithmetic units: it can learn arbitrary power functions for
positive, negative, and small numbers. We showed that the NPU outperforms its
main competitor (NALU) and reaches performance that is on par with the
multiplication specialist NMU (Sec.~\ref{sub:simple_arithmetic_task} \&
\ref{sub:large_scale_arithmetic_task}).\\
Additionally we have demonstrated that the NPU converges consistently, even on
sequential tasks and that can be used as a highly interpretable by recovering
the  governing equations of dynamical systems purely from data
(Sec.~\ref{sub:fractional_sir_model}).

@@

\newpage
\appendix
\section*{Appendix}%
\label{sec:appendix}

@@

\begin{table}[h]
  \centering
  \caption{Model definitions for the fSIR task.}
  \label{tab:fsir_models}
  \begin{tabular}{llll}
    \toprule
    Model & Layer 1 & Layer 2 & Layer 3 \\
    \midrule
    NPU & $\npu(3,\,h)$ & $\nau(h,\,3)$ & -- \\
    NPU & NPU$_{\text{real}}(3,\,h)$ & $\nau(h,\,3)$ & -- \\
    Dense & $\dense(2,\,h,\sigma)$ & $\dense(h,\,h,\,\sigma)$ & $\dense(h,\,3)$ \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[h]
  \centering
  \caption{Testing error on the simple arithmetic task for the different models
  (i.e. mean of each heatmap in Fig.~\ref{fig:simple_err}). Each value is
  obtained by averaging the error of 20 models.}
  \label{tab:simple_err}
  \input{../plots/simple_err.tex}
\end{table}

\begin{table}[h]
  \centering
  \caption{Model definitions for the simple arithmetic task.}
  \label{tab:models_simple_task}
  \begin{tabular}{llll}
    \toprule
    Model & Layer 1 & Layer 2 & Layer 3 \\
    \midrule
    NPU & NAU(2,\,6) & NPU(6,\,2) & -- \\
    NMU & NAU(2,\,6) & NMU(6,\,2) & -- \\
    NALU & NALU(2,\,6) & NALU(6,\,2) & -- \\
    iNALU & iNALU(2,\,6) & iNALU(6,\,2) & -- \\
    Dense & Dense(2,\,10,\,$\sigma$) & Dense(10,\,10,\,$\sigma$) & Dense(10,\,2) \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[h]
  \centering
  \caption{Model definitions for the large scale arithmetic task.}
  \label{tab:models_large_task}
  \begin{tabular}{llll}
    \toprule
    Model & Layer 1 & Layer 2\\
    \midrule
    NPU & NAU(100,\,100) & NPU(100,\,1) \\
    NPU & NAU(100,\,100) & NPU(100,\,1) \\
    NMU & NAU(100,\,100) & NMU(100,\,1) \\
    NALU & NALU(100,\,100) & NALU(100,\,1) \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{table}[h]
  \centering
  \caption{Training parameters for the large scale arithmetic task. The
  $\beta$-parameters define the stepwise exponential growth of the L1
  regularization with start, step, growth, and end.}
  \label{tab:training_large_scale}
  \begin{tabular}{lcccccc}
    \toprule
    Task & Learning rate & Iterations & $\beta_{\text{start}}$ &
      $\beta_{\text{end}}$ & $\beta_{\text{step}}$  & $\beta_{\text{growth}}$ \\
    \midrule
    Add  & 1e-2 & 1e5 & 1e-5 & 1e-4 & 10\,000 & 10 \\
    Mult & 5e-3 & 1e5 & 1e-5 & 1e-7 & 10\,000 & 10 \\
    Div  & 5e-3 & 1e5 & 1e-9 & 1e-7 & 10\,000 & 10 \\
    Sqrt & 5e-3 & 1e5 & 1e-6 & 1e-4 & 10\,000 & 10 \\
    \bottomrule
  \end{tabular}
\end{table}
